---
title: "Normalization strategy comparison for Data-driven analysis of isobarically labeled proteomic data."
author: "Joris Van Houtven, Piotr Prostko"
date: '`r format(Sys.time(), "%B %d, %Y,%H:%M")`'
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
    theme: flatly
    code_folding: "hide"
editor_options: 
  chunk_output_type: console 
params:
  input_data_p: 'input_data_msstatstmt.Rds'
  suffix_p: 'msstatstmt'
  load_outputdata_p: FALSE
  subsample_p: 0   
---
  
```{r, setup, include=FALSE}
# default knitr options for all chunks
knitr::opts_chunk$set(
  message=FALSE,
  warning=FALSE,
  fig.width=12,
  fig.height=8
)
```

```{r}
library(ggplot2)
library(stringi)
library(gridExtra)
library(dendextend)
library(kableExtra)
library(psych)
library(limma)
library(tidyverse)
library(preprocessCore)
library(CONSTANd)  # install from source: https://github.com/PDiracDelta/CONSTANd/
library(NOMAD)  # devtools::install_github("carlmurie/NOMAD")
source('other_functions.R')
source('plotting_functions.R')
```

```{r}
data.list <- readRDS(params$input_data_p)
dat.l <- data.list$dat.l # data in long format
# which proteins were spiked in?
spiked.proteins <- dat.l %>% distinct(Protein) %>% filter(stri_detect(Protein, fixed='ups')) %>% pull %>% as.character
tmp=dat.l %>% distinct(Protein) %>% pull %>% as.character
# protein subsampling
if (params$subsample_p>0 & params$subsample_p==floor(params$subsample_p) & params$subsample_p<=length(tmp)){
  sub.prot <- tmp[sample(1:length(tmp), size=params$subsample_p)]
  if (length(spiked.proteins)>0) sub.prot <- c(sub.prot,spiked.proteins)
  dat.l <- dat.l %>% filter(Protein %in% sub.prot)
}
```

```{r}
# specify # of varying component variants and their names
variant.names <- c('median_sweeping', 'CONSTANd', 'NOMAD', 'quantile1')
n.comp.variants <- length(variant.names)
scale.vec <- 'log'
# pick reference condition for making plots / doing DEA
referenceCondition <- '0.5'
# specify colours corresponding to biological conditions
condition.color <- tribble(
  ~Condition, ~Color,
  "0.125", 'black',
  "0.5", 'blue',
  "0.667", 'green',
  "1", 'red' )
# create data frame with sample info (distinct Run,Channel, Sample, Condition, Color)
sample.info <- get_sample_info(dat.l, condition.color)
channelNames <- remove_factors(unique(sample.info$Channel))
```

# Unit component: log2 transformation of reporter ion intensities

```{r}
dat.unit.l <- dat.l %>% mutate(response=log2(intensity)) %>% select(-intensity)
```

# Normalization component

```{r}
# switch to wide format
dat.unit.w <- pivot_wider(data = dat.unit.l, id_cols=-one_of(c('Condition', 'BioReplicate')), names_from=Channel, values_from=response)
```

```{r}
dat.norm.w <- emptyList(variant.names)
```

## median sweeping (1)

```{r}
if (!load_outputdata){
# subtract the spectrum median log2intensity from the observed log2intensities
dat.norm.w$median_sweeping <- dat.unit.w
dat.norm.w$median_sweeping[,channelNames] <- median_sweep(dat.norm.w$median_sweeping[,channelNames], 1, '-') 
}
```

## CONSTANd

```{r}
if (!load_outputdata){
# dat.unit.l entries are in long format so all have same colnames and no channelNames
x.split <- split(dat.unit.w, dat.unit.w$Run)  # apply CONSTANd to each Run separately
x.split.norm  <- lapply(x.split, function(y) {
  y[,channelNames] <- CONSTANd(y[,channelNames])$normalized_data
  return(y)
})
dat.norm.w$CONSTANd <- bind_rows(x.split.norm)
}
```

## NOMAD

We apply NOMAD on the PSM level instead of the peptide level.

```{r}
if (!load_outputdata){
# doRobust=F: use means, like CONSTANd; doLog=F: values are already transformed.
dat.nomadnorm <- nomadNormalization(dat.unit.l$response, dat.unit.l %>% rename(iTRAQ=Channel) %>% as.data.frame, doRobust = FALSE, doiTRAQCorrection = FALSE, doLog = FALSE)
dat.nomadnorm$x$response <- dat.nomadnorm$y
dat.norm.w$NOMAD <- pivot_wider(data = dat.nomadnorm$x, id_cols=-one_of(c('Condition', 'BioReplicate')), names_from=iTRAQ, values_from=response)
}
```

## Quantile (1)

We apply quantile normalization to each Run separately, and then re-scale the observations so that the mean observation within in each run is set equal to the mean observation across all runs.

```{r}
if (!load_outputdata){
grand_average <- mean(as.matrix(dat.unit.w[,channelNames]))
x.split <- split(dat.unit.w, dat.unit.w$Run)  
x.split.norm <- lapply(x.split, function(y) {
  # apply normalizeQuantiles to each Run separately
  y[,channelNames] <- normalize.quantiles(as.matrix(y[,channelNames]))
  # make averages of all runs equal.
  y[,channelNames] <- y[,channelNames] / mean(colMeans(y[,channelNames])) * grand_average
  return(y)
})
dat.norm.w$quantile1 <- bind_rows(x.split.norm)
}
```

# Summarization component: Median summarization

Summarize quantification values from PSM to peptide (first step) to protein (second step).

```{r}
if (!load_outputdata){
# normalized data
dat.norm.summ.w <- lapply(dat.norm.w, function(x){
  y <- x %>% group_by(Run, Protein, Peptide) %>% summarise_at(.vars = channelNames, .funs = median, na.rm=T) %>% summarise_at(.vars = channelNames, .funs = median, na.rm=T) %>% ungroup()
  return(y)})
}
```

Let's also summarize the non-normalized data for comparison in the next section.
```{r}
if (!load_outputdata){
# non-normalized data
# group by (run,)protein,peptide then summarize twice (once on each level)
# add select() statement because summarise_at is going bananas over character columns
dat.nonnorm.summ.w <- dat.unit.w %>% group_by(Run, Protein, Peptide) %>% select(Run, Protein, Peptide, channelNames) %>% summarise_at(.vars = channelNames, .funs = median) %>% select(Run, Protein, channelNames) %>% summarise_at(.vars = channelNames, .funs = median) %>% ungroup()
}
```

```{r}
if (!load_outputdata){
# make data completely wide (also across runs)
## normalized data
dat.norm.summ.w2 <- lapply(dat.norm.summ.w, function(x) {
  return(x %>% pivot_wider(names_from = Run, values_from = all_of(channelNames), names_glue = "{Run}:{.value}"))
})
}
```

# Normalization component (2)

## Median sweeping (2)

```{r}
if (!load_outputdata){
# median sweeping: in each channel, subtract median computed across all proteins within the channel, separately for each MS run.
tmp <- dat.norm.summ.w2$median_sweeping %>% select(-Protein)
dat.norm.summ.w2$median_sweeping[, colnames(dat.norm.summ.w2$median_sweeping)!='Protein'] <- median_sweep(tmp, 2, '-') 
}
```

## Quantile (2)

```{r}
if (!load_outputdata){
# apply normalizeQuantiles again, now to the data from all runs simultaneously
dat.norm.summ.w2$quantile1[,sample.info$Sample] <- normalize.quantiles(as.matrix(dat.norm.summ.w2$quantile1[,sample.info$Sample]))
}
```

```{r, echo=FALSE}
if (!load_outputdata){
# propagate the second round of median sweeping and quantile1 normalization from dat.norm.summ.w2 to dat.norm.summ.w
  dat.norm.summ.w$median_sweeping <- dat.norm.summ.w2$median_sweeping %>% pivot_longer(cols=-one_of('Protein'), names_to='Sample', values_to='response', values_drop_na = T) %>% inner_join(sample.info[,c('Sample', 'Run', 'Channel', 'Condition')], by='Sample') %>% pivot_wider(id_cols=-one_of('Condition', 'Sample'), names_from=Channel, values_from=response)
  
  dat.norm.summ.w$quantile1 <- dat.norm.summ.w2$quantile1 %>% pivot_longer(cols=-one_of('Protein'), names_to='Sample', values_to='response', values_drop_na = T) %>% inner_join(sample.info[,c('Sample', 'Run', 'Channel', 'Condition')], by='Sample') %>% pivot_wider(id_cols=-one_of('Condition', 'Sample'), names_from=Channel, values_from=response)
}
```

## Quantile norm of protein-level data by condition

```{r}
if (!load_outputdata){
dat.nonnorm.summ.l <- to_long_format(dat.nonnorm.summ.w, sample.info)
x.split <- split(dat.nonnorm.summ.l, dat.nonnorm.summ.l$Condition)
x.split.norm <- lapply(x.split, function(x){
  tmp.wide <- pivot_wider(data=x, id_cols='Protein', names_from=Sample, values_from=response) 
  tmp.wide[, colnames(tmp.wide)!='Protein'] <- normalize.quantiles(as.matrix(tmp.wide %>% select(-Protein)))
  tmp <- pivot_longer(tmp.wide, cols=-one_of('Protein'), names_to='Sample', values_to='response') %>% drop_na() %>% inner_join(sample.info[,c('Sample', 'Run', 'Channel', 'Condition')], by='Sample')
  return(tmp)
})
x <- bind_rows(x.split.norm)
dat.norm.summ.w$quantile2 <- pivot_wider(data = x, id_cols=-one_of('Condition', 'Sample'), names_from=Channel, values_from=response)
dat.norm.summ.w2$quantile2 <- dat.norm.summ.w$quantile2 %>% pivot_wider(names_from = Run, values_from = all_of(channelNames), names_glue = "{Run}:{.value}")
}
variant.names <- c('median_sweeping', 'CONSTANd', 'NOMAD', 'quantile1', 'quantile2')
n.comp.variants <- length(variant.names)
```

# QC plots

```{r}
if (!load_outputdata){
# make data completely wide (also across runs)
## non-normalized data
dat.nonnorm.summ.w2 <- dat.nonnorm.summ.w %>% pivot_wider(names_from = Run, values_from = all_of(channelNames), names_glue = "{Run}:{.value}")
}
```

```{r, echo=FALSE}
if (load_outputdata) load(paste0('datadriven_normalization_outdata_', params$suffix_p, '.rda'))
```

## Boxplots

```{r}
# use (half-)wide format
par(mfrow=c(3,2))
boxplot_w(dat.nonnorm.summ.w,sample.info, 'raw')
for (i in 1:n.comp.variants){
  boxplot_w(dat.norm.summ.w[[variant.names[i]]], sample.info, paste('normalized', variant.names[i], sep='_'))}
```

## MA plots

MA plots of two single samples taken from condition 1 and condition 0.125, measured in different MS runs (samples *Mixture2_1:127C* and *Mixture1_2:129N*, respectively).

```{r}
# use wide2 format
p <- vector('list', n.comp.variants+1)
p[[1]] <- maplot_ils(dat.nonnorm.summ.w2, 'Mixture2_1:127C', 'Mixture1_2:129N', scale.vec, 'raw', spiked.proteins)
for (i in 1: n.comp.variants){
 p[[i+1]]<- maplot_ils(dat.norm.summ.w2[[variant.names[i]]], 'Mixture2_1:127C', 'Mixture1_2:129N', scale.vec, paste('normalized', variant.names[i], sep='_'), spiked.proteins)}
grid.arrange(grobs = p, ncol=2, nrow=3)
```

MA plots of all samples from condition `1` and condition `0.125` (quantification values averaged within condition).

```{r}
channels.num <- sample.info %>% filter(Condition=='1') %>% select(Sample) %>% pull
channels.denom <- sample.info %>% filter(Condition=='0.125') %>% select(Sample) %>% pull
p <- vector('list', n.comp.variants+1)
p[[1]] <- maplot_ils(dat.nonnorm.summ.w2, channels.num, channels.denom, scale.vec, 'raw', spiked.proteins)
for (i in 1: n.comp.variants){
 p[[i+1]]<- maplot_ils(dat.norm.summ.w2[[variant.names[i]]], channels.num, channels.denom, scale.vec, paste('normalized', variant.names[i], sep='_'), spiked.proteins)}
grid.arrange(grobs = p, ncol=2)
```

```{r, echo=FALSE}
#dat.nonnorm.summ.l <- to_long_format(dat.nonnorm.summ.w, sample.info)
dat.norm.summ.l <- lapply(dat.norm.summ.w, to_long_format, sample.info)
```

## PCA plots

### Using all proteins
```{r}
par(mfrow=c(3,2))
pcaplot_ils(dat.nonnorm.summ.w2 %>% select(-'Protein'), info=sample.info, 'raw')
for (i in 1:n.comp.variants){
pcaplot_ils(dat.norm.summ.w2[[variant.names[i]]] %>% select(-'Protein'), info=sample.info, paste('normalized', variant.names[i], sep='_'))}
```

### Using spiked proteins only
```{r}
par(mfrow=c(3,2))
pcaplot_ils(dat.nonnorm.summ.w2 %>% filter(Protein %in% spiked.proteins) %>% select(-'Protein'), info=sample.info, 'raw')
for (i in 1:n.comp.variants){
  pcaplot_ils(dat.norm.summ.w2[[variant.names[i]]] %>% filter(Protein %in% spiked.proteins) %>% select(-'Protein'), info=sample.info, paste('normalized', variant.names[i], sep='_'))}
```

## HC (hierarchical clustering) plots

### Using all proteins

```{r, fig.width=12, fig.height=15}
par(mfrow=c(3,2))
dendrogram_ils(dat.nonnorm.summ.w2 %>% select(-Protein), info=sample.info, 'raw')
for (i in 1:n.comp.variants){
  dendrogram_ils(dat.norm.summ.w2[[variant.names[i]]] %>% select(-Protein), info=sample.info, paste('normalized', variant.names[i], sep='_'))}
```

## Run effect p-value plot

Below the histograms of p-values from the linear model explaining the response variable with Run as a covariate.

```{r}
dat <- vector('list',length(dat.norm.summ.l)+1)
dat[[1]] <- dat.nonnorm.summ.l; dat[2:length(dat)] <- dat.norm.summ.l[1:length(dat.norm.summ.l)]
names(dat) <- c('raw', names(dat.norm.summ.l))
run_effect_plot(dat)
```

# DEA component: Moderated t-test

TODO:
- Also try to log-transform the intensity case, to see if there are large differences in the t-test results.
  - done. remove this code?
NOTE:
- actually, lmFit (used in moderated_ttest) was built for log2-transformed data. However, supplying untransformed intensities can also work. This just means that the effects in the linear model are also additive on the untransformed scale, whereas for log-transformed data they are multiplicative on the untransformed scale. Also, there may be a bias which occurs from biased estimates of the population means in the t-tests, as mean(X) is not equal to exp(mean(log(X))).
```{r}
if (!load_outputdata){
design.matrix <- get_design_matrix(referenceCondition, sample.info)
dat.dea <- emptyList(names(dat.norm.summ.w2))
for (i in 1:n.comp.variants) {
  this_scale <- scale.vec
  d <- column_to_rownames(as.data.frame(dat.norm.summ.w2[[variant.names[i]]]), 'Protein')
  dat.dea[[variant.names[i]]] <- moderated_ttest(dat=d, design.matrix, scale=this_scale)}
# also see what the unnormalized results would look like
n.comp.variants <- n.comp.variants + 1
variant.names <- c(variant.names, 'raw')
dat.dea$raw <- moderated_ttest(dat=column_to_rownames(dat.nonnorm.summ.w2, 'Protein'), design.matrix, scale='log')
}
```

```{r, echo=FALSE}
# save output data
if (!load_outputdata){
save(dat.nonnorm.summ.l
     ,dat.norm.summ.l
     ,dat.nonnorm.summ.w
     ,dat.norm.summ.w
     ,dat.nonnorm.summ.w2
     ,dat.norm.summ.w2
     ,dat.dea, file=paste0('datadriven_normalization_outdata_', params$suffix_p, '.rda'))
}
```

# Results comparison

## Confusion matrix

```{r, results='asis'}
cm <- conf_mat(dat.dea, 'q.mod', 0.05, spiked.proteins)
print_conf_mat(cm, referenceCondition)
```

## Scatter plots

```{r}
# character vectors containing logFC and p-values columns
dea.cols <- colnames(dat.dea[[1]])
logFC.cols <- dea.cols[stri_detect_fixed(dea.cols, 'logFC')]
significance.cols <- dea.cols[stri_detect_fixed(dea.cols, 'q.mod')]
n.contrasts <- length(logFC.cols)

scatterplot_ils(dat.dea, significance.cols, 'q-values', spiked.proteins, referenceCondition)
scatterplot_ils(dat.dea, logFC.cols, 'log2FC', spiked.proteins, referenceCondition)
```

## Volcano plots

```{r}
for (i in 1:n.contrasts){
  volcanoplot_ils(dat.dea, i, spiked.proteins, referenceCondition)}
```

## Violin plots

Let's see whether the spiked protein fold changes make sense
```{r}
# plot theoretical value (horizontal lines) and violin per variant
violinplot_ils(lapply(dat.dea, function(x) x[spiked.proteins, logFC.cols]), referenceCondition)
```

# Conclusions

# Session information

```{r}
sessionInfo()
```