---
title: "DEA strategy comparison for Data-driven analysis of isobarically labeled proteomic data."
author: "Joris Van Houtven, Piotr Prostko"
date: '`r format(Sys.time(), "%B %d, %Y,%H:%M")`'
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
    theme: flatly
    code_folding: "hide"
editor_options: 
  chunk_output_type: console
---
  
```{r, setup, include=FALSE}
# default knitr options for all chunks
knitr::opts_chunk$set(
  message=FALSE,
  warning=FALSE,
  fig.width=12,
  fig.height=7
)
```

```{r}
library(ggplot2)
library(stringi)
library(gridExtra)
library(dendextend)
library(kableExtra)
library(limma)
library(psych)
library(tidyverse)
library(CONSTANd)  # install from source: https://github.com/PDiracDelta/CONSTANd/
library(matrixTests)
library(coin)
library(ROTS)
```

This notebook presents isobaric labeling data analysis strategy that includes data-driven normalization. 

We will check how varying analysis components [summarization/normalization/differential abundance testing methods] changes end results of a quantitative proteomic study.

```{r}
source('./other_functions.R')
source('./plotting_functions.R')

# you should either make a symbolic link in this directory
data.list <- readRDS('input_data.rds')
dat.l <- data.list$dat.l # data in long format
# dat.w <- data.list$dat.w # data in wide format
if ('X' %in% colnames(dat.l)) { dat.l$X <- NULL }

# remove shared peptides
shared.peptides <- dat.l %>% filter(!shared.peptide)

# keep spectra with isolation interference <30 and no missing quantification channels
dat.l <- dat.l %>% filter(isoInterOk & noNAs)

# which proteins were spiked in?
spiked.proteins <- dat.l %>% distinct(Protein) %>% filter(stri_detect(Protein, fixed='ups')) %>% pull %>% as.character

# which peptides were identified in each MS run?
unique.pep=dat.l %>% 
  group_by(Run) %>%
  distinct(Peptide) %>% 
  mutate(val=1)
unique.pep <- xtabs(val~Peptide+Run, data=unique.pep)
tmp <- apply(unique.pep, 1, function(x) all(x==1))
inner.peptides <- rownames(unique.pep)[tmp]

#TEMP -remove!
# tmp=dat.l %>% distinct(Protein) %>% pull %>% as.character
# dat.l <- dat.l %>% filter(Protein %in% c(tmp[sample(1:length(tmp), size=20)], spiked.proteins))
```

```{r}
# specify # of varying component variants and their names
variant.names <- c('LIMMA', 'Wilcoxon', 'permutation_test', 'ROTS')
n.comp.variants <- length(variant.names)
scale.vec <- c('log', 'log', 'log', 'log') 
# pick reference channel and condition for making plots / doing DEA
referenceChannel <- '127C'
referenceCondition <- '0.5'
# specify colours corresponding to biological conditions
condition.colour <- tribble(
  ~Condition, ~Colour,
  "0.125", 'black',
  "0.5", 'blue',
  "0.667", 'green',
  "1", 'red' )
# create data frame with sample info (distinct Run,Channel, Sample, Condition, Colour)
sample.info <- get_sample_info(dat.l, condition.colour)
channelNames <- remove_factors(unique(sample.info$Channel))
# set seed for permutation test
seed=9107
```

# Unit component: log2 transformation of reporter ion intensities

```{r}
dat.unit.l <- dat.l
```

```{r}
dat.unit.l <- dat.l %>% mutate(response=log2(Intensity)) %>% select(-Intensity)
grand_median <- median(dat.unit.l$response, na.rm=TRUE)
```

# Normalization component: medianSweeping (1)

```{r}
# switch to wide format
dat.unit.w <- pivot_wider(data = dat.unit.l, id_cols=-one_of(c('Condition', 'BioReplicate')), names_from=Channel, values_from=response) 
```

```{r}
# subtract the spectrum median log2intensity from the observed log2intensities
dat.norm.w <- dat.unit.w
dat.norm.w[,channelNames] <- sweep(dat.norm.w[,channelNames], 1, apply(dat.norm.w[,channelNames], 1, median) )
```

# Summarization component: Median summarization

Summarize quantification values from PSM to peptide (first step) to protein (second step).

```{r}
# normalized data
# group by (run,)protein,peptide then summarize twice (once on each level)
dat.norm.summ.w <- dat.norm.w %>% group_by(Run, Protein, Peptide) %>% summarise_at(.vars = channelNames, .funs = median) %>% summarise_at(.vars = channelNames, .funs = median) %>% ungroup() 
```

Notice that the row sums are not equal to Ncols anymore, because the median summarization does not preserve them (but mean summarization does).

Let's also summarize the non-normalized data for comparison in the next section.
```{r}
# non-normalized data
# add select() statement because summarise_at is going bananas over character columns
dat.nonnorm.summ.w <- dat.unit.w %>% group_by(Run, Protein, Peptide) %>% select(Run, Protein, Peptide, channelNames) %>% summarise_at(.vars = channelNames, .funs = median) %>% select(Run, Protein, channelNames) %>% summarise_at(.vars = channelNames, .funs = median) %>% ungroup()
```

# Normalization component: medianSweeping (2)

```{r}
# medianSweeping: in each channel, subtract median computed across all proteins within the channel
# do the above separately for each MS run
x.split <- split(dat.norm.summ.w, dat.norm.summ.w$Run)  
x.split.norm  <- lapply(x.split, function(y) {
  y[,channelNames] <- sweep(y[,channelNames], 2, apply(y[,channelNames], 2, median) )
  return(y)
})
dat.norm.summ.w <- bind_rows(x.split.norm)
```

# QC plots

```{r}
# make data completely wide (also across runs)

## non-normalized data
dat.nonnorm.summ.w2 <- dat.nonnorm.summ.w %>% pivot_wider(names_from = Run, values_from = all_of(channelNames), names_glue = "{Run}:{.value}")

## normalized data
dat.norm.summ.w2 <- dat.norm.summ.w %>% pivot_wider(names_from = Run, values_from = all_of(channelNames), names_glue = "{Run}:{.value}")
```

## Boxplots

```{r}
# use (half-)wide format
par(mfrow=c(1,2))
  boxplot_w(dat.nonnorm.summ.w,sample.info, 'Raw')
  boxplot_w(dat.norm.summ.w, sample.info, 'Normalized')
par(mfrow=c(1,1))
```

## MA plots

MA plots of two single samples taken from condition 1 and condition 0.125, measured in different MS runs (samples *Mixture2_1:127C* and *Mixture1_2:129N*, respectively).

```{r}
# different unit variants require different computation of fold changes and average abuandance: additive or multiplicative scale; see maplot_ils function 
# use wide2 format
p1 <- maplot_ils(dat.nonnorm.summ.w2, 'Mixture2_1:127C', 'Mixture1_2:129N', scale='log', 'Raw', spiked.proteins)
p2 <- maplot_ils(dat.norm.summ.w2, 'Mixture2_1:127C', 'Mixture1_2:129N', scale='log', 'Normalized', spiked.proteins)
grid.arrange(p1, p2, ncol=2)  
```

## MA plots of all samples from condition `1` and condition `0.125` (quantification values averaged within condition).

```{r}
# different unit variants require different computation of fold changes and average abuandance: additive or multiplicative scale; see maplot_ils function 
channels.num <- sample.info %>% filter(Condition=='1') %>% select(Sample) %>% pull
channels.denom <- sample.info %>% filter(Condition=='0.125') %>% select(Sample) %>% pull

p1 <- maplot_ils(dat.nonnorm.summ.w2, channels.num, channels.denom, scale='log', 'Raw', spiked.proteins)
p2 <- maplot_ils(dat.norm.summ.w2, channels.num, channels.denom, scale='log', 'Normalized', spiked.proteins)
grid.arrange(p1, p2, ncol=2)  
```

## CV (coefficient of variation) plots

```{r}
dat.nonnorm.summ.w$Mixture <- unlist(lapply(stri_split(dat.nonnorm.summ.w$Run,fixed='_'), function(x) x[1]))
dat.nonnorm.summ.l <- to_long_format(dat.nonnorm.summ.w, sample.info)

dat.norm.summ.w$Mixture <- unlist(lapply(stri_split(dat.norm.summ.w$Run,fixed='_'), function(x) x[1]))
dat.norm.summ.l <- to_long_format(dat.norm.summ.w, sample.info)

par(mfrow=c(1, 2))
add.constant=grand_median
cvplot_ils(dat=dat.nonnorm.summ.l, feature.group='Protein', xaxis.group='Condition', title='Raw')
cvplot_ils(dat=dat.norm.summ.l, feature.group='Protein', xaxis.group='Condition', title='Normalized', add.constant=grand_median)
par(mfrow=c(1, 1))  
```

## PCA plots

### Using all proteins
```{r}
par(mfrow=c(1, 2))
  pcaplot_ils(dat.nonnorm.summ.w2 %>% select(-'Protein'), info=sample.info, 'Raw')
  pcaplot_ils(dat.norm.summ.w2 %>% select(-'Protein'), info=sample.info, 'Normalized')
par(mfrow=c(1, 1))  
```

### Using spiked proteins only
```{r}
par(mfrow=c(1, 2))
  pcaplot_ils(dat.nonnorm.summ.w2 %>% filter(Protein %in% spiked.proteins) %>% select(-'Protein'), info=sample.info, 'Raw')
  pcaplot_ils(dat.norm.summ.w2 %>% filter(Protein %in% spiked.proteins) %>% select(-'Protein'), info=sample.info, 'Normalized')
par(mfrow=c(1, 1))  
```

## HC (hierarchical clustering) plots

### Using all proteins

```{r}
par(mfrow=c(1, 2))
  dendrogram_ils(dat.nonnorm.summ.w2 %>% select(-Protein), info=sample.info, 'Raw')
  dendrogram_ils(dat.norm.summ.w2 %>% select(-Protein), info=sample.info, 'Normalized')
par(mfrow=c(1, 1))  
```

# DEA component

## Moderated t-test

TODO:
- Also try to log-transform the intensity case, to see if there are large differences in the t-test results.
  - done. remove this code?
NOTE:
- actually, lmFit (used in moderated_ttest) was built for log2-transformed data. However, supplying untransformed intensities can also work. This just means that the effects in the linear model are also additive on the untransformed scale, whereas for log-transformed data they are multiplicative on the untransformed scale. Also, there may be a bias which occurs from biased estimates of the population means in the t-tests, as mean(X) is not equal to exp(mean(log(X))).
```{r}
design.matrix <- get_design_matrix(referenceCondition, sample.info)
dat.dea <- emptyList(variant.names)
this_scale <- scale.vec[1]
d <- column_to_rownames(as.data.frame(dat.norm.summ.w2), 'Protein')
dat.dea$LIMMA <- moderated_ttest(dat=d, design.matrix, scale=this_scale) 
```

## Wilcoxon test

```{r}
otherConditions <- dat.l %>% distinct(Condition) %>% pull(Condition) %>% as.character %>% sort
otherConditions <- otherConditions[-match(referenceCondition, otherConditions)]
dat.dea$Wilcoxon <- wilcoxon_test(dat.norm.summ.w2, sample.info, referenceCondition, otherConditions, logFC.method='ratio')
```

## Permutation test

```{r}
dat.dea$permutation_test <- permutation_test(dat.norm.summ.l, referenceCondition, otherConditions, seed=seed, distribution='exact')
```

## ROTS

```{r}
dat.dea$ROTS <- rots_test(dat.norm.summ.w2, referenceCondition, otherConditions)
```

# Results comparison

## Confusion matrix

```{r, results='asis'}
cm <- conf_mat(dat.dea, 'q.mod', 0.05, spiked.proteins)
print_conf_mat(cm, referenceCondition)
```

## Scatter plots

```{r}
# character vectors containing logFC and p-values columns
dea.cols <- colnames(dat.dea[[1]])
logFC.cols <- dea.cols[stri_detect_fixed(dea.cols, 'logFC')]
significance.cols <- dea.cols[stri_detect_fixed(dea.cols, 'q.mod')]
n.contrasts <- length(logFC.cols)

scatterplot_ils(dat.dea, significance.cols, 'q-values', spiked.proteins)
scatterplot_ils(dat.dea, logFC.cols, 'log2FC', spiked.proteins)
```

## Volcano plots

```{r}
for (i in 1:n.contrasts){
  volcanoplot_ils(dat.dea, i, spiked.proteins) }
```

## Violin plots

Let's see whether the spiked protein fold changes make sense
```{r}
# plot theoretical value (horizontal lines) and violin per variant
violinplot_ils(lapply(dat.dea, function(x) x[spiked.proteins, logFC.cols]), referenceCondition)
```

# Conclusions

# Session information

```{r}
sessionInfo()
```