---
title: "Differential Expression Analysis strategy comparison for Data-driven analysis of isobarically labeled proteomic data."
author: "Joris Van Houtven, Piotr Prostko"
date: '`r format(Sys.time(), "%B %d, %Y,%H:%M")`'
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
    theme: flatly
    code_folding: "hide"
editor_options: 
  chunk_output_type: console
params:
  input_data_p: 'input_data_msstatstmt.Rds'
  suffix_p: 'msstatstmt'
  load_outputdata_p: FALSE
  subsample_p: 0
---
  
```{r, setup, include=FALSE}
# default knitr options for all chunks
knitr::opts_chunk$set(
  message=FALSE,
  warning=FALSE,
  fig.width=12,
  fig.height=7
)
```

```{r}
library(ggplot2)
library(stringi)
library(gridExtra)
library(dendextend)
library(kableExtra)
library(limma)
library(psych)
library(tidyverse)
library(CONSTANd)  # install from source: https://github.com/PDiracDelta/CONSTANd/
library(matrixTests)
library(coin)
library(ROTS)
source('other_functions.R')
source('plotting_functions.R')
```

```{r}
data.list <- readRDS(params$input_data_p)
dat.l <- data.list$dat.l # data in long format
# which proteins were spiked in?
spiked.proteins <- dat.l %>% distinct(Protein) %>% filter(stri_detect(Protein, fixed='ups')) %>% pull %>% as.character
tmp=dat.l %>% distinct(Protein) %>% pull %>% as.character
# protein subsampling
if (params$subsample_p>0 & params$subsample_p==floor(params$subsample_p) & params$subsample_p<=length(tmp)){
  sub.prot <- tmp[sample(1:length(tmp), size=params$subsample_p)]
  if (length(spiked.proteins)>0) sub.prot <- c(sub.prot,spiked.proteins)
  dat.l <- dat.l %>% filter(Protein %in% sub.prot)
}
```

```{r}
# specify # of varying component variants and their names
variant.names <- c('moderated_ttest', 'Wilcoxon', 'permutation_test', 'ROTS')
n.comp.variants <- length(variant.names)
scale.vec <- 'log'
# set seed for permutation test
seed=9107
# pick reference condition for making plots / doing DEA
referenceCondition <- '0.5'
# specify colours corresponding to biological conditions
condition.color <- tribble(
  ~Condition, ~Color,
  "0.125", 'black',
  "0.5", 'blue',
  "0.667", 'green',
  "1", 'red' )
# create data frame with sample info (distinct Run,Channel, Sample, Condition, Color)
sample.info <- get_sample_info(dat.l, condition.color)
channelNames <- remove_factors(unique(sample.info$Channel))
# specify samples and conditions for ma plots
ma.onesample.num <- 'Mixture2_1:127C'
ma.onesample.denom <- 'Mixture1_2:129N'
ma.allsamples.num <- '1'
ma.allsamples.denom <- '0.125'
```

# Unit scale component: log2 transformation of reporter ion intensities

```{r}
dat.unit.l <- dat.l
```

```{r}
dat.unit.l <- dat.l %>% mutate(response=log2(intensity)) %>% select(-intensity)
```

# Normalization component: medianSweeping (1)

```{r}
# switch to wide format
dat.unit.w <- pivot_wider(data = dat.unit.l, id_cols=-one_of(c('Condition', 'BioReplicate')), names_from=Channel, values_from=response) 
```

```{r}
# subtract the spectrum median log2intensity from the observed log2intensities
dat.norm.w <- dat.unit.w
dat.norm.w[,channelNames] <- median_sweep(dat.norm.w[,channelNames], 1, '-')
```

# Summarization component: Median summarization

Summarize quantification values from PSM to peptide (first step) to protein (second step).

```{r, eval=!params$load_outputdata_p}
# normalized data
dat.norm.summ.w <- dat.norm.w %>% group_by(Run, Protein, Peptide) %>% summarise_at(.vars = channelNames, .funs = median) %>% summarise_at(.vars = channelNames, .funs = median) %>% ungroup() 
```

Let's also summarize the non-normalized data for comparison in the next section.
```{r, eval=!params$load_outputdata_p}
# non-normalized data
# group by (run,)protein,peptide then summarize twice (once on each level)
# add select() statement because summarise_at is going bananas over character columns
dat.nonnorm.summ.w <- dat.unit.w %>% group_by(Run, Protein, Peptide) %>% select(Run, Protein, Peptide, channelNames) %>% summarise_at(.vars = channelNames, .funs = median) %>% select(Run, Protein, channelNames) %>% summarise_at(.vars = channelNames, .funs = median) %>% ungroup()
```

# Normalization component: medianSweeping (2)

```{r, eval=!params$load_outputdata_p}
# medianSweeping: in each channel, subtract median computed across all proteins within the channel
# do the above separately for each MS run
x.split <- split(dat.norm.summ.w, dat.norm.summ.w$Run)  
x.split.norm  <- lapply(x.split, function(y) {
  y[,channelNames] <- median_sweep(y[,channelNames], 2, '-')
  return(y)
})
dat.norm.summ.w <- bind_rows(x.split.norm)
```

# QC plots

```{r, eval=!params$load_outputdata_p}
# make data completely wide (also across runs)
## normalized data
dat.norm.summ.w2 <- dat.norm.summ.w %>% pivot_wider(names_from = Run, values_from = all_of(channelNames), names_glue = "{Run}:{.value}")

## non-normalized data
dat.nonnorm.summ.w2 <- dat.nonnorm.summ.w %>% pivot_wider(names_from = Run, values_from = all_of(channelNames), names_glue = "{Run}:{.value}")
```

```{r, echo=FALSE, eval=params$load_outputdata_p}
load(paste0('datadriven_DEA_outdata_', params$suffix_p, '.rda'))
```

## Boxplots

```{r}
# use (half-)wide format
par(mfrow=c(1,2))
boxplot_w(dat.nonnorm.summ.w,sample.info, 'raw')
boxplot_w(dat.norm.summ.w, sample.info, 'normalized')
```

## MA plots

MA plots of two single samples taken from condition `r ma.allsamples.num` and condition `r ma.allsamples.denom`, measured in different MS runs (samples *`r ma.onesample.num`* and *`r ma.onesample.denom`*, respectively).

```{r}
# use wide2 format
p1 <- maplot_ils(dat.nonnorm.summ.w2, ma.onesample.num, ma.onesample.denom, scale='log', 'raw', spiked.proteins)
p2 <- maplot_ils(dat.norm.summ.w2, ma.onesample.num, ma.onesample.denom, scale='log', 'normalized', spiked.proteins)
grid.arrange(p1, p2, ncol=2)  
```

MA plots of all samples from condition *`r ma.allsamples.num`* and condition *`r ma.allsamples.denom`* (quantification values averaged within condition).

```{r}
channels.num <- sample.info %>% filter(Condition==ma.allsamples.num) %>% select(Sample) %>% pull
channels.denom <- sample.info %>% filter(Condition==ma.allsamples.denom) %>% select(Sample) %>% pull
p1 <- maplot_ils(dat.nonnorm.summ.w2, channels.num, channels.denom, scale='log', 'raw', spiked.proteins)
p2 <- maplot_ils(dat.norm.summ.w2, channels.num, channels.denom, scale='log', 'normalized', spiked.proteins)
grid.arrange(p1, p2, ncol=2)  
```

```{r, echo=FALSE, eval=!params$load_outputdata_p}
dat.nonnorm.summ.l <- to_long_format(dat.nonnorm.summ.w, sample.info)
dat.norm.summ.l <- to_long_format(dat.norm.summ.w, sample.info)
```

## PCA plots

### Using all proteins
```{r}
par(mfrow=c(1, 2))
pcaplot_ils(dat.nonnorm.summ.w2 %>% select(-'Protein'), info=sample.info, 'raw')
pcaplot_ils(dat.norm.summ.w2 %>% select(-'Protein'), info=sample.info, 'normalized')
```

### Using spiked proteins only (if applicable)
```{r, eval=length(spiked.proteins)>0}
par(mfrow=c(1, 2))
pcaplot_ils(dat.nonnorm.summ.w2 %>% filter(Protein %in% spiked.proteins) %>% select(-'Protein'), info=sample.info, 'raw')
pcaplot_ils(dat.norm.summ.w2 %>% filter(Protein %in% spiked.proteins) %>% select(-'Protein'), info=sample.info, 'normalized')
```

## HC (hierarchical clustering) plots

### Using all proteins

```{r}
par(mfrow=c(1, 2))
dendrogram_ils(dat.nonnorm.summ.w2 %>% select(-Protein), info=sample.info, 'raw')
dendrogram_ils(dat.norm.summ.w2 %>% select(-Protein), info=sample.info, 'normalized')
```

## Run effect p-value plot

Below the histograms of p-values from the linear model explaining the response variable with Run as a covariate.

```{r}
dat <- list(dat.nonnorm.summ.l,dat.norm.summ.l)
names(dat) <- c('raw','normalized')
run_effect_plot(dat)
```

# DEA component

## Moderated t-test

TODO:
- Also try to log-transform the intensity case, to see if there are large differences in the t-test results.
  - done. remove this code?
NOTE:
- actually, lmFit (used in moderated_ttest) was built for log2-transformed data. However, supplying untransformed intensities can also work. This just means that the effects in the linear model are also additive on the untransformed scale, whereas for log-transformed data they are multiplicative on the untransformed scale. Also, there may be a bias which occurs from biased estimates of the population means in the t-tests, as mean(X) is not equal to exp(mean(log(X))).
```{r, eval=!params$load_outputdata_p}
design.matrix <- get_design_matrix(referenceCondition, sample.info)
dat.dea <- emptyList(variant.names)
this_scale <- scale.vec
d <- column_to_rownames(as.data.frame(dat.norm.summ.w2), 'Protein')
dat.dea$moderated_ttest <- moderated_ttest(dat=d, design.matrix, scale=this_scale) 
```

## Wilcoxon test

```{r, eval=!params$load_outputdata_p}
otherConditions <- dat.l %>% distinct(Condition) %>% pull(Condition) %>% as.character %>% sort
otherConditions <- otherConditions[-match(referenceCondition, otherConditions)]
dat.dea$Wilcoxon <- wilcoxon_test(dat.norm.summ.w2, sample.info, referenceCondition, otherConditions, logFC.method='ratio')
```

## Permutation test

```{r, eval=!params$load_outputdata_p}
dat.dea$permutation_test <- permutation_test(dat.norm.summ.l, referenceCondition, otherConditions, seed=seed, distribution='exact')
```

## ROTS

```{r, eval=!params$load_outputdata_p}
dat.dea$ROTS <- rots_test(dat.norm.summ.w2, sample.info, referenceCondition, otherConditions)
```

```{r, echo=FALSE, eval=!params$load_outputdata_p}
# save output data
save(dat.nonnorm.summ.l
     ,dat.norm.summ.l
     ,dat.nonnorm.summ.w
     ,dat.norm.summ.w
     ,dat.nonnorm.summ.w2
     ,dat.norm.summ.w2
     ,dat.dea, file=paste0('datadriven_DEA_outdata_', params$suffix_p, '.rda'))
```

# Results comparison

## Confusion matrix

```{r, results='asis'}
cm <- conf_mat(dat.dea, 'q.mod', 0.05, spiked.proteins)
print_conf_mat(cm, referenceCondition)
```

## Scatter plots

```{r}
# character vectors containing logFC and p-values columns
dea.cols <- colnames(dat.dea[[1]])
logFC.cols <- dea.cols[stri_detect_fixed(dea.cols, 'logFC')]
significance.cols <- dea.cols[stri_detect_fixed(dea.cols, 'q.mod')]
n.contrasts <- length(logFC.cols)

scatterplot_ils(dat.dea, significance.cols, 'q-values', spiked.proteins, referenceCondition)
scatterplot_ils(dat.dea[c('moderated_ttest', 'Wilcoxon')], logFC.cols, 'log2FC', spiked.proteins, referenceCondition)
```

## Volcano plots

```{r}
for (i in 1:n.contrasts){
  volcanoplot_ils(dat.dea, i, spiked.proteins, referenceCondition)}
```

## Violin plots

Let's see whether the spiked protein fold changes make sense
```{r, eval=length(spiked.proteins)>0}
# plot theoretical value (horizontal lines) and violin per variant
violinplot_ils(lapply(dat.dea[c('moderated_ttest', 'Wilcoxon')], function(x) x[spiked.proteins, logFC.cols]), referenceCondition)
```

# Conclusions

# Session information

```{r}
sessionInfo()
```