---
title: "Data-driven analysis of isobaric labels data. DEA component."
author: "Joris Van Houtven"
date: '`r format(Sys.time(), "%B %d, %Y,%H:%M")`'
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
    theme: flatly
    code_folding: "hide"
editor_options: 
  chunk_output_type: console
---
  
```{r, setup, include=FALSE}
# default knitr options for all chunks
knitr::opts_chunk$set(
  message=FALSE,
  warning=FALSE,
  fig.width=12,
  fig.height=7
)
```

```{r}
library(ggplot2)
library(stringi)
library(gridExtra)
library(dendextend)
library(kableExtra)
library(limma)
library(psych)
library(tidyverse)
library(CONSTANd)  # install from source: https://github.com/PDiracDelta/CONSTANd/
library(matrixTests)
```

This notebook presents isobaric labeling data analysis strategy that includes data-driven normalization. 

We will check how varying analysis components [summarization/normalization/differential abundance testing methods] changes end results of a quantitative proteomic study.

```{r}
source('./other_functions.R')
source('./plotting_functions.R')

# you should either make a symbolic link in this directory
data.list <- readRDS('input_data.rds')
dat.l <- data.list$dat.l # data in long format
# dat.w <- data.list$dat.w # data in wide format
if ('X' %in% colnames(dat.l)) { dat.l$X <- NULL }

# remove shared peptides
shared.peptides <- dat.l %>% filter(!shared.peptide)

# keep spectra with isolation interference <30 and no missing quantification channels
dat.l <- dat.l %>% filter(isoInterOk & noNAs)

# which proteins were spiked in?
spiked.proteins <- dat.l %>% distinct(Protein) %>% filter(stri_detect(Protein, fixed='ups')) %>% pull %>% as.character

# which peptides were identified in each MS run?
unique.pep=dat.l %>% 
  group_by(Run) %>%
  distinct(Peptide) %>% 
  mutate(val=1)
unique.pep <- xtabs(val~Peptide+Run, data=unique.pep)
tmp <- apply(unique.pep, 1, function(x) all(x==1))
inner.peptides <- rownames(unique.pep)[tmp]
```

```{r}
# specify # of varying component variants and their names
n.comp.variants <- 3
variant.names <- c('LIMMA', 'Wilcoxon', 'Permutation test')
scale.vec <- c('log', 'log', 'log') 
# pick reference channel and condition for making plots / doing DEA
referenceChannel <- '127C'
referenceCondition <- '0.5'
```

# Unit component

```{r}
# do not use list (see modelbased notebook comment)
dat.unit.l <-list()
```

## log2 transformation of reporter ion intensities

```{r}
dat.unit.l[[1]] <- dat.l %>% mutate(response=log2(Intensity)) %>% select(-Intensity)
```

# Normalization component

```{r}
# switch to wide format
dat.unit.w <- lapply(dat.unit.l, function(x) {
  pivot_wider(data = x, id_cols=-one_of(c('Condition', 'BioReplicate')), names_from=Channel, values_from=response) } )
```

## CONSTANd

```{r}
# dat.unit.l entries are in long format so all have same colnames and no quanCols
quanCols = setdiff(colnames(dat.unit.w[[1]]), colnames(dat.unit.l[[1]]))
dat.norm.w <- lapply(dat.unit.w, function(x) {
  x.split <- split(x, x$Run)  # apply CONSTANd to each Run separately
  x.split.norm  <- lapply(x.split, function(y) {
    y[,quanCols] <- CONSTANd(y[,quanCols])$normalized_data
    return(y) } )
  return(bind_rows(x.split.norm)) } )
```

# Summarization component

Summarize quantification values from PSM to peptide (first step) to protein (second step).

## Median summarization (PSM to peptide to protein)

```{r}
# normalized data
study.design=read.delim('msstatstmt_studydesign.csv')
dat.norm.summ.w <- lapply(dat.norm.w, function(x) {
  # group by (run,)protein,peptide then summarize twice (once on each level)
  return( x %>% group_by(Run, Protein, Peptide) %>% summarise_at(.vars = quanCols, .funs = median) %>% summarise_at(.vars = quanCols, .funs = median) %>% ungroup() ) } )
```

Notice that the row sums are not equal to Ncols anymore, because the median summarization
does not preserve them (but mean summarization does).

# QC plots

```{r}
# make data completely wide (also across runs)

## normalized data
dat.norm.summ.w2 <- lapply(dat.norm.summ.w, function(x) {
  return( x %>% pivot_wider(names_from = Run, values_from = all_of(quanCols)) ) } )

# make vectors with condition labels and color coding corresponding to samples in wide format data
colors.condition <- tribble(
  ~Condition, ~Col,
  "0.125", 'black',
  "0.5", 'blue',
  "0.667", 'green',
  "1", 'red'
)
run_channel_condition <- expand_grid(Channel=quanCols, Run=unique(study.design$Run)) %>% left_join(study.design, by=c('Channel', 'Run')) %>% select(Run, Channel, Condition)
colors.condition.map <- run_channel_condition %>% unite(Channel, Channel:Run) %>% left_join(colors.condition, by='Condition')
ord <- match(colnames(dat.norm.summ.w2[[1]]), colors.condition.map$Channel)
ord <- ord[!is.na(ord)]  # drop first entry which is NA

# important: these two vectors contain colors and condition labels corresponding to data in wide2 format
cols.vec <- colors.condition.map[ord, 'Col'] %>% pull
conditions.vec <- colors.condition.map[ord, 'Condition'] %>% pull
```

# DEA component

## Moderated t-test

TODO:
- Also try to log-transform the intensity case, to see if there are large differences in the t-test results.
  - done. remove this code?
NOTE:
- actually, lmFit (used in moderated_ttest) was built for log2-transformed data. However, supplying untransformed intensities can also work. This just means that the effects in the linear model are also additive on the untransformed scale, whereas for log-transformed data they are multiplicative on the untransformed scale. Also, there may be a bias which occurs from biased estimates of the population means in the t-tests, as mean(X) is not equal to exp(mean(log(X))).
```{r}
design.matrix <- get_design_matrix(referenceCondition, study.design)
dat.dea <- emptyList(variant.names)
# when not using lists of length 1, this for-loop disappears.
for (i in seq_along(dat.norm.summ.w2)) {
  this_scale <- scale.vec[match(names(dat.dea)[i], variant.names)]
  d <- column_to_rownames(as.data.frame(dat.norm.summ.w2[[i]]), 'Protein')
  dat.dea[[i]] <- moderated_ttest(dat=d, design.matrix, scale=this_scale) }
```

## Wilcoxon test

```{r}
otherConditions <- dat.l %>% distinct(Condition) %>% pull(Condition) %>% as.character %>% sort
otherConditions <- otherConditions[-match(referenceCondition, otherConditions)]
# please use names
dat.dea$wilcoxon <- wilcoxon_test(dat.norm.summ.w2[[1]], referenceCondition, otherConditions)
```

## Permutation test

```{r}
dat.dea[[3]] <- permutation_test(dat.norm.summ.w2[[1]], referenceCondition, otherConditions, B=1000, seed=394)
```

# Results comparison

## Confusion matrix

```{r, results='asis'}
cm <- conf.mat(dat.dea, 'q.mod', 0.05, spiked.proteins)
print.conf.mat(cm)
```

## Scatter plots

```{r}
# character vectors containing logFC and p-values columns
dea.cols <- colnames(dat.dea[[1]])
logFC.cols <- dea.cols[stri_detect_fixed(dea.cols, 'logFC')]
q.cols <- dea.cols[stri_detect_fixed(dea.cols, 'q.mod')]
n.contrasts <- length(logFC.cols)

scatterplot.ils(dat.dea, q.cols, 'p-values')
scatterplot.ils(dat.dea, logFC.cols, 'log2FC')
```

## Volcano plots

```{r}
for (i in 1:n.contrasts){
  volcanoplot.ils(dat.dea, i, spiked.proteins) }
```

## Violin plots

Let's see whether the spiked protein fold changes make sense
```{r}
# plot theoretical value (horizontal lines) and violin per condition
dat.spiked.logfc <- lapply(dat.dea, function(x) x[spiked.proteins,logFC.cols])
dat.spiked.logfc.l <- lapply(dat.spiked.logfc, function(x) {
  x %>% rename_with(function(y) sapply(y, function(z) strsplit(z, '_')[[1]][2])) %>% pivot_longer(cols = everything(), names_to = 'condition', values_to = 'logFC') %>% add_column(Protein=rep(rownames(x), each=length(colnames(x)))) })
violinplot.ils(lapply(dat.spiked.logfc.l, filter, condition != referenceCondition))
```

# Conclusions

# Session information

```{r}
sessionInfo()
```