---
title: "Model-based analysis of isobaric labels data. Unit component."
author: "Piotr Prostko"
date: '`r format(Sys.time(), "%B %d, %Y,%H:%M")`'
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
    theme: flatly
editor_options: 
  chunk_output_type: console
---

```{r, setup, include=FALSE}
# default knitr options for all chunks
knitr::opts_chunk$set(
  message=FALSE,
  warning=FALSE,
  echo=FALSE
)
```


```{r}
library(caret)
library(DEP)
library(nlme)
library(lme4)
library(lmerTest)
library(limma)
library(ggplot2)
library(stringi)
library(RColorBrewer)
library(gridExtra)
library(tidyverse)
```

This notebook presents isobaric labeling data analysis strategy that includes model-based normalization [data-driven normalization]. 

We will check how varying analysis units [summarization/normalization/differential abundance testing methods] changes end results of a quantitative proteomic study.

```{r}
source('./other_functions.r')
source('./plotting_functions.r')

path.data <- 'G:/My Drive/Isobaric labeling strategies/data'
data.list <- readRDS(paste0(path.data, '/input_data.rds'))
dat.l <- data.list$dat.l # data in long format
dat.w <- data.list$dat.w # data in wide format

# keep spectra with isolation interference <30 and no missing quantification channels
dat.l <- dat.l %>% filter(isoInterOk=='Y' & noNAs=='Y')

# which proteins were spiked in?
spiked.proteins <- dat.l %>% distinct(Protein) %>% filter(stri_detect(Protein, fixed='ups')) %>% pull

# which peptides were identified in each MS run?
unique.pep=dat.l %>% 
  group_by(Run) %>%
  distinct(Peptide) %>% 
  mutate(val=1)
unique.pep <- xtabs(val~Peptide+Run, data=unique.pep)
tmp <- apply(unique.pep, 1, function(x) all(x==1))
inner.peptides <- rownames(unique.pep)[tmp]
```

```{r}
# specify # of varying component variants and their names
n.components <- 3
variant.names <- c('log2Intensity', 'Intensity', 'log2Ratio')
```

# Unit component

```{r}
comp1 <- vector('list', n.components)
names(comp1) <- variant.names
```

## variant 1: log2 transformation of reporter ion intensities

```{r}
comp1[[1]] <- dat.l %>% mutate(response=log2(Intensity)) %>% select(-Intensity)
```

## variant 2: original scale (not log-transformed) of reporter ion intensities
```{r}
comp1[[2]] <- dat.l %>% rename(response=Intensity)
```

## variant 3: log2-transformation of intensity ratios (channel 127C in denominator)

```{r}
# or we can try another channel as reference
denom.df <- dat.l %>% filter(Channel=='127C') %>% rename(denom=Intensity) %>%
  select(Run, Protein, Peptide, RT, Charge, PTM, denom)
comp1[[3]] <- dat.l %>% left_join(denom.df, by=c('Run', 'Protein', 'Peptide', 'RT', 'Charge', 'PTM')) %>% mutate(response=Intensity/denom) %>% select(-c(Intensity))
```

# Summarization component

```{r}
comp2 <- comp1
```

## variant 1: no summarization

<!-- ## variant 2: PSM to peptide summarization -->

# Normalization component

## variant1: mixed model spec 1

```{r}
comp3 <- lapply(comp2, function(x) {
  mod <- lmer(response ~ Run + Run:Channel +
                 (1|Protein)  + (1|Peptide), data=x)
  x$response <- residuals(mod)
  return(x)
})
```

<!-- ## variant2: mixed model spec 2 -->

<!-- ## variant3: mixed model spec 3 -->

## Normalization diagnostic plots

```{r}
# PSM data needs to be aggregated prior to PCA plots and HC plots (they requires features in the intersection of all MS runs - this is not possible for PSM data). In order to be consisent, other normalization plots will be based on the aggregated data
dat.l.summplot <- lapply(comp2, function(x) aggFunc(x, 'response', 'mean')) # before normalization
dat.l.normplot <- lapply(comp3, function(x) aggFunc(x, 'response', 'mean')) # after normalization

# now create data sets in wide format
# before normalization
dat.w.summplot <- lapply(dat.l.summplot, function(x) {
  dat.tmp <- unclass(xtabs('response~ Peptide + I(Run:Channel)', data=x, drop.unused.levels = TRUE))
  return(dat.tmp)
})
# after normalization
dat.w.normplot <- lapply(dat.l.normplot, function(x) {
  dat.tmp <- unclass(xtabs('response~ Peptide + I(Run:Channel)', data=x, drop.unused.levels = TRUE))
  return(dat.tmp)
})

# make vectors with condition labels and color coding corresponding to samples in wide format data
colors.condition <- tribble(
  ~Condition, ~Col,
  "0.125", 'black',
  "0.5", 'blue',
  "0.667", 'green',
  "1", 'red'
)
# sufficient to do that only on the first items of dat.l.normplot and dat.w.normplot lists
colors.condition.map <- dat.l.normplot[[1]] %>% distinct(Run:Channel, Condition) %>% left_join(colors.condition, by='Condition')
ord <- match(colnames(dat.w.normplot[[1]]), colors.condition.map$`Run:Channel`)

# important: these two vectors contain colors and condition labels corresponding to data in wide format
cols.vec <- colors.condition.map[ord, 'Col']  %>% pull
conditions.vec <- colors.condition.map[ord, 'Condition']  %>% pull
```

Boxplots:
```{r}
par(mfrow=c(3,2))
for (i in 1:n.components){
  boxplot(response~Run:Channel, dat.l.summplot[[i]], main=paste('Before normalization', variant.names[i], sep='_'), notch=TRUE)
  boxplot(response~Run:Channel, dat.l.normplot[[i]], main=paste('After normalization', variant.names[i], sep='_'), notch=TRUE)
}
par(mfrow=c(1,1))  
```

MA plots:

CV (coefficient of variation) plots:

PCA plots:

HC (hierarchical clustering) plots:

# DEA component

```{r}
shared.peptides <- dat.l %>% filter(shared.peptide=='Y') %>% 
  distinct(Peptide) %>% pull
```

## variant1: Mixed model (intra-protein correlation) + eBayes

```{r}
comp4 <- lapply(comp3, function(x){
  tmp <- x %>% filter(!(Peptide %in% shared.peptides)) # remove shared peptides before DEA
  out <- mixed.model.dea(dat=tmp, mod.formula='response ~ Condition + (1|Run:Channel)', conditions=c('1', '0.125'))
  return(out)
})

```

<!-- ## variant2: DEqMS -->

## Comparison of differential abundance testing

# Conclusions










