---
title: "Summarization strategy comparison for Model-based analysis of isobarically labeled proteomic data."
author: "Piotr Prostko, Joris Van Houtven"
date: '`r format(Sys.time(), "%B %d, %Y,%H:%M")`'
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
    theme: flatly
    code_folding: "hide"
editor_options: 
  chunk_output_type: console
params:
  input_data_p: 'input_data_msstatstmt.Rds'
  suffix_p: 'msstatstmt'
  load_outputdata_p: FALSE
  subsample_p: 0
---

```{r, setup, include=FALSE}
# default knitr options for all chunks
knitr::opts_chunk$set(
  message=FALSE,
  warning=FALSE,
  fig.width=12,
  fig.height=7
)
```

```{r}
library(caret)
library(lme4)
library(lmerTest)
library(ggplot2)
library(stringi)
library(gridExtra)
library(ggfortify)
library(dendextend)
library(psych)
library(kableExtra)
library(tidyverse)
library(dtplyr)
source('other_functions.R')
source('plotting_functions.R')
```


```{r}
data.list <- readRDS(params$input_data_p)
dat.l <- data.list$dat.l # data in long format
# which proteins were spiked in?
spiked.proteins <- dat.l %>% distinct(Protein) %>% filter(stri_detect(Protein, fixed='ups')) %>% pull %>% as.character
tmp=dat.l %>% distinct(Protein) %>% pull %>% as.character
# protein subsampling
if (params$subsample_p>0 & params$subsample_p==floor(params$subsample_p) & params$subsample_p<=length(tmp)){
  sub.prot <- tmp[sample(1:length(tmp), size=params$subsample_p)]
  if (length(spiked.proteins)>0) sub.prot <- c(sub.prot,spiked.proteins)
  dat.l <- dat.l %>% filter(Protein %in% sub.prot)
}
```

```{r}
# specify # of varying component variants and their names
variant.names <- c('no_summ', 'median', 'sum')
n.comp.variants <- length(variant.names)
# pick reference condition for making plots / doing DEA
referenceCondition <- '0.5'
# specify colours corresponding to biological conditions
condition.color <- tribble(
  ~Condition, ~Color,
  "0.125", 'black',
  "0.5", 'blue',
  "0.667", 'green',
  "1", 'red' )
# create data frame with sample info (distinct Run,Channel, Sample, Condition, Color)
sample.info <- get_sample_info(dat.l, condition.color)
channelNames <- remove_factors(unique(sample.info$Channel))
# specify samples and conditions for ma plots
ma.onesample.num <- 'Mixture2_1:127C'
ma.onesample.denom <- 'Mixture1_2:129N'
ma.allsamples.num <- '1'
ma.allsamples.denom <- '0.125'
```

# Unit scale component: log2 transformation of reporter ion intensities

```{r}
dat.unit.l <- dat.l %>% mutate(response=log2(intensity)) %>% select(-intensity)
```

# Summarization component

```{r}
dat.summ.l <- emptyList(variant.names)
```

## no summarization

```{r}
dat.summ.l$no_summ <- dat.unit.l
```

## PSM to peptide summarization using median

```{r, eval=!params$load_outputdata_p}
dat.summ.l$median <- aggFunc(dat.unit.l, 'response', group.vars=c('Mixture', 'TechRepMixture', 'Run', 'Channel', 'Condition', 'BioReplicate', 'Protein',  'Peptide'), 'median')
```

## PSM to peptide summarization using sum

```{r, eval=!params$load_outputdata_p}
dat.summ.l$sum <- aggFunc(dat.unit.l, 'response', group.vars=c('Mixture', 'TechRepMixture', 'Run', 'Channel', 'Condition', 'BioReplicate', 'Protein',  'Peptide'), 'sum')
```

# Normalization component: linear mixed model

```{r, eval=!params$load_outputdata_p}
dat.norm.l <- lapply(dat.summ.l, function(x) {
  mod <- lmer(response ~ Run + Run:Channel + (1|Protein)  + (1|Run:Peptide), data=x)
  x$response <- residuals(mod)+fixef(mod)['(Intercept)']
  return(x)})
```

# QC plots

```{r, eval=!params$load_outputdata_p}
dat.nonnorm.summ.l <- lapply(dat.summ.l, function(x) aggFunc(x, 'response', group.vars=c('Mixture', 'TechRepMixture', 'Run', 'Channel', 'Condition', 'BioReplicate', 'Protein', 'Peptide'), 'median')) # before normalization (you still need to aggregate the data because of 'dat.summ.l <- dat.unit.l' operation)
dat.nonnorm.summ.l <- lapply(dat.nonnorm.summ.l, function(x) aggFunc(x, 'response', group.vars=c('Mixture', 'TechRepMixture', 'Run', 'Channel', 'Condition', 'BioReplicate', 'Protein'), 'median'))
dat.norm.summ.l <- lapply(dat.norm.l, function(x) aggFunc(x, 'response', group.vars=c('Mixture', 'TechRepMixture', 'Run', 'Channel', 'Condition', 'BioReplicate', 'Protein', 'Peptide'), 'median')) # after normalization
dat.norm.summ.l <- lapply(dat.norm.summ.l, function(x) aggFunc(x, 'response', group.vars=c('Mixture', 'TechRepMixture', 'Run', 'Channel', 'Condition', 'BioReplicate', 'Protein'), 'median')) # after normalization

# make data completely wide (also across runs)
## normalized data
dat.norm.summ.w2 <- lapply(dat.norm.summ.l, function(x) {
  dat.tmp <- pivot_wider(data=x, id_cols=Protein, names_from=Run:Channel, values_from=response, names_sep=':')
  return(dat.tmp) } )

## non-normalized data
dat.nonnorm.summ.w2 <- lapply(dat.nonnorm.summ.l, function(x) {
  dat.tmp <- pivot_wider(data=x, id_cols=Protein, names_from=Run:Channel, values_from=response, names_sep=':')
  return(dat.tmp) } )
```

```{r, echo=FALSE, eval=params$load_outputdata_p}
load(paste0('modelbased_summarization_outdata_', params$suffix_p, '.rda'))
```

## Boxplots

```{r}
par(mfrow=c(2,2))
for (i in 1:n.comp.variants){
  boxplot_ils(dat.nonnorm.summ.l[[variant.names[i]]], paste('raw', variant.names[i], sep='_'))
  boxplot_ils(dat.norm.summ.l[[variant.names[i]]], paste('normalized', variant.names[i], sep='_'))}
```

## MA plots

MA plots of two single samples taken from condition `r ma.allsamples.num` and condition `r ma.allsamples.denom`, measured in different MS runs (samples *`r ma.onesample.num`* and *`r ma.onesample.denom`*, respectively).

```{r}
for (i in 1:n.comp.variants){
  p1 <- maplot_ils(dat.nonnorm.summ.w2[[variant.names[i]]], ma.onesample.num, ma.onesample.denom, scale='log', paste('raw', variant.names[i], sep='_'), spiked.proteins)
  p2 <- maplot_ils(dat.norm.summ.w2[[variant.names[i]]], ma.onesample.num, ma.onesample.denom, scale='log', paste('normalized', variant.names[i], sep='_'), spiked.proteins)
  grid.arrange(p1, p2, ncol=2)}
```

MA plots of all samples from condition *`r ma.allsamples.num`* and condition *`r ma.allsamples.denom`* (quantification values averaged within condition).

```{r}
channels.num <- sample.info %>% filter(Condition==ma.allsamples.num) %>% distinct(Sample) %>% pull
channels.denom <- sample.info %>% filter(Condition==ma.allsamples.denom) %>% distinct(Sample) %>% pull
for (i in 1:n.comp.variants){
  p1 <- maplot_ils(dat.nonnorm.summ.w2[[variant.names[i]]], channels.num, channels.denom, scale='log', paste('raw', variant.names[i], sep='_'), spiked.proteins)
  p2 <- maplot_ils(dat.norm.summ.w2[[variant.names[i]]], channels.num, channels.denom, scale='log', paste('normalized', variant.names[i], sep='_'), spiked.proteins)
  grid.arrange(p1, p2, ncol=2)}
```

## PCA plots

### Using all proteins

```{r}
par(mfrow=c(1, 2))
for (i in 1:n.comp.variants){
  pcaplot_ils(dat.nonnorm.summ.w2[[variant.names[i]]] %>% select(-'Protein'), info=sample.info, paste('raw', variant.names[i], sep='_'))
  pcaplot_ils(dat.norm.summ.w2[[variant.names[i]]] %>% select(-'Protein'), info=sample.info, paste('normalized', variant.names[i], sep='_'))}
par(mfrow=c(1, 1))
```

### Using spiked proteins only (if applicable)

```{r, eval=length(spiked.proteins)>0}
par(mfrow=c(1, 2))
for (i in 1:n.comp.variants){
  pcaplot_ils(dat.nonnorm.summ.w2[[variant.names[i]]] %>% filter(Protein %in% spiked.proteins) %>% select(-'Protein'), info=sample.info, paste('raw', variant.names[i], sep='_'))
  pcaplot_ils(dat.norm.summ.w2[[variant.names[i]]] %>% filter(Protein %in% spiked.proteins) %>% select(-'Protein'), info=sample.info, paste('normalized', variant.names[i], sep='_'))}
par(mfrow=c(1, 1))
```

## HC (hierarchical clustering) plots

### Using all proteins

```{r}
par(mfrow=c(1, 2))
for (i in 1:n.comp.variants){
  dendrogram_ils(dat.nonnorm.summ.w2[[variant.names[i]]] %>% select(-'Protein'), info=sample.info, paste('raw', variant.names[i], sep='_'))
  dendrogram_ils(dat.norm.summ.w2[[variant.names[i]]] %>% select(-'Protein'), info=sample.info, paste('normalized', variant.names[i], sep='_'))}
par(mfrow=c(1, 1))
```

## Run effect p-value plot

Below the histograms of p-values from the linear model explaining the response variable with Run as a covariate.

```{r}
plots <- vector('list', n.comp.variants)
for (i in 1:n.comp.variants){
dat <- list(dat.nonnorm.summ.l[[variant.names[i]]], dat.norm.summ.l[[variant.names[i]]])
names(dat) <- c(paste('raw', variant.names[i], sep='_'), paste('normalized', variant.names[i], sep='_'))
plots[[i]] <- run_effect_plot(dat)}
grid.arrange(grobs = plots, nrow=n.comp.variants) 
```

# DEA component: linear mixed model

Intra-protein correlation modelled with 1|Run:Channel random effect.

```{r, eval=!params$load_outputdata_p}
dat.dea <- lapply(dat.norm.l, function(x){
  return(lmm_dea(dat=x, mod.formula='response ~ Condition + (1|Run:Channel)', referenceCondition, scale='log'))})
```

```{r}
# character vectors containing logFC and p-values columns
dea.cols <- colnames(dat.dea[[1]])
logFC.cols <- dea.cols[stri_detect_fixed(dea.cols, 'logFC')]
significance.cols <- dea.cols[stri_detect_fixed(dea.cols, 'q.mod')]
n.contrasts <- length(logFC.cols)
```

```{r, echo=FALSE, eval=!params$load_outputdata_p}
# save output data
save(dat.nonnorm.summ.l
     ,dat.norm.summ.l
     ,dat.nonnorm.summ.w2
     ,dat.norm.summ.w2
     ,dat.norm.l
     ,dat.summ.l
     ,dat.dea, file=paste0('modelbased_summarization_outdata_', params$suffix_p, '.rda'))
```

# Results comparison

## Confusion matrix

```{r, results='asis'}
cm <- conf_mat(dat.dea, 'q.mod', 0.05, spiked.proteins)
print_conf_mat(cm, referenceCondition)
```

## Scatter plots

```{r}
scatterplot_ils(dat.dea, significance.cols, 'q-values', spiked.proteins, referenceCondition)
scatterplot_ils(dat.dea, logFC.cols, 'log2FC', spiked.proteins, referenceCondition)
```

## Volcano plots

```{r}
for (i in 1:n.contrasts){
  volcanoplot_ils(dat.dea, i, spiked.proteins, referenceCondition)}
```

## Violin plots

Let's see whether the spiked protein fold changes make sense
```{r, eval=length(spiked.proteins)>0}
# plot theoretical value (horizontal lines) and violin per variant
violinplot_ils(lapply(dat.dea, function(x) x[spiked.proteins, logFC.cols]), referenceCondition)
```

# Conclusions

# Session information
```{r}
sessionInfo()
```
