---
title: "Model-based analysis of isobaric labels data. Summarization component."
author: "Piotr Prostko"
date: '`r format(Sys.time(), "%B %d, %Y,%H:%M")`'
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
    theme: flatly
editor_options: 
  chunk_output_type: console
---

```{r, setup, include=FALSE}
# default knitr options for all chunks
knitr::opts_chunk$set(
  message=FALSE,
  warning=FALSE,
  echo=FALSE
)
```

```{r}
library(caret)
# library(DEP)
# library(nlme)
library(lme4)
library(lmerTest)
library(limma)
library(ggplot2)
library(stringi)
library(RColorBrewer)
library(gridExtra)
library(tidyverse)
```

This notebook presents isobaric labeling data analysis strategy that includes model-based normalization [data-driven normalization]. 

We will check how varying summarization methods [analysis unit/normalization/differential abundance testing methods] changes end results of a quantitative proteomic study.

```{r}
source('./other_functions.r')
source('./plotting_functions.r')

path.data <- 'G:/My Drive/Isobaric labeling strategies/data'
data.list <- readRDS(paste0(path.data, '/input_data.rds'))
dat.l <- data.list$dat.l # data in long format
dat.w <- data.list$dat.w # data in wide format

# keep spectra with isolation interference <30 and no missing quantification channels
dat.l <- dat.l %>% filter(isoInterOk=='Y' & noNAs=='Y')

# which proteins were spiked in?
spiked.proteins <- dat.l %>% distinct(Protein) %>% filter(stri_detect(Protein, fixed='ups')) %>% pull

# which peptides were identified in each MS run?
unique.pep=dat.l %>% 
  group_by(Run) %>%
  distinct(Peptide) %>% 
  mutate(val=1)
unique.pep <- xtabs(val~Peptide+Run, data=unique.pep)
tmp <- apply(unique.pep, 1, function(x) all(x==1))
inner.peptides <- rownames(unique.pep)[tmp]
```

```{r}
# specify # of varying component variants and their names
n.comp.variants <- 3
variant.names <- c('no_summ', 'median_summ', 'sum_summ')
```

# Unit component

```{r}
dat.unit.l <- vector('list', 1)
```

## log2 transformation of reporter ion intensities

```{r}
dat.unit.l[[1]] <- dat.l %>% mutate(response=log2(Intensity)) %>% select(-Intensity)
```

<!-- ## original scale (not log-transformed) of reporter ion intensities -->

<!-- ## log2-transformation of intensity ratios (channel 127C in denominator) -->

# Summarization component

```{r}
dat.summ.l <- vector('list', n.comp.variants)
names(dat.summ.l) <- variant.names
```

## no summarization

```{r}
dat.summ.l[[1]] <- dat.unit.l[[1]]
```

## PSM to peptide summarization using median

```{r}
dat.summ.l[[2]] <- aggFunc(dat.unit.l[[1]], 'response', agg.method='median')
```

## PSM to peptide summarization using sum

```{r}
dat.summ.l[[3]] <- aggFunc(dat.unit.l[[1]], 'response', agg.method='sum')
```

# Normalization component

## mixed model spec 1

```{r}
dat.norm.l <- lapply(dat.summ.l, function(x) {
  mod <- lmer(response ~ Run + Run:Channel +
                 (1|Protein)  + (1|Peptide), data=x)
  x$response <- residuals(mod)
  return(x)
})
```

<!-- ## variant2: mixed model spec 2 -->

<!-- ## variant3: mixed model spec 3 -->

# Normalization diagnostic plots

```{r}
# PSM data needs to be aggregated prior to PCA plots and HC plots (they require features in the intersection of all MS runs - this is not possible for PSM data). In order to be consisent, other normalization plots will be based on the aggregated data

# before normalization
dat.summplot.l <- dat.summ.l
dat.summplot.l[[1]] <- aggFunc(dat.summ.l[[1]], 'response', 'mean') # only first variant requires aggregation to peptide level

# after normalization
dat.normplot.l <- dat.norm.l
dat.normplot.l[[1]] <- aggFunc(dat.norm.l[[1]], 'response', 'mean') # only first variant requires aggregation to peptide level

# now create data sets in wide format
# before normalization
dat.summplot.w <- lapply(dat.summplot.l, function(x) {
  dat.tmp <- unclass(xtabs('response~ Peptide + I(Run:Channel)', data=x, drop.unused.levels = TRUE))
  return(dat.tmp)
})

# after normalization
dat.normplot.w <- lapply(dat.normplot.l, function(x) {
  dat.tmp <- unclass(xtabs('response~ Peptide + I(Run:Channel)', data=x, drop.unused.levels = TRUE))
  return(dat.tmp)
})

# make vectors with condition labels and color coding corresponding to samples in wide format data
colors.condition <- tribble(
  ~Condition, ~Col,
  "0.125", 'black',
  "0.5", 'blue',
  "0.667", 'green',
  "1", 'red'
)
# sufficient to do that only on the first items of dat.normplot.l and dat.normplot.w lists
colors.condition.map <- dat.normplot.l[[1]] %>% distinct(Run:Channel, Condition) %>% left_join(colors.condition, by='Condition')
ord <- match(colnames(dat.normplot.w[[1]]), colors.condition.map$`Run:Channel`)

# important: these two vectors contain colors and condition labels corresponding to data in wide format
cols.vec <- colors.condition.map[ord, 'Col']  %>% pull
conditions.vec <- colors.condition.map[ord, 'Condition']  %>% pull
```

Boxplots:

MA plots:

CV (coefficient of variation) plots:

PCA plots:

HC (hierarchical clustering) plots:

# DEA component

```{r}
shared.peptides <- dat.l %>% filter(shared.peptide=='Y') %>% 
  distinct(Peptide) %>% pull
```

## mixed model (intra-protein correlation) + eBayes

```{r}
dat.dea <- lapply(dat.norm.l, function(x){
  tmp <- x %>% filter(!(Peptide %in% shared.peptides)) # remove shared peptides before DEA
  out <- mixed.model.dea(dat=tmp[1:1000,], mod.formula='response ~ Condition + (1|Run:Channel)', conditions=c('1', '0.125')) # subsetting to first 1000 records to speed up knitting
  return(out)
})
```

<!-- ## DEqMS -->

# Results comparison

# Conclusions

# Session information
```{r}
sessionInfo()
```









